\textsf{\textbf{A Toturial of BioConductor R packages}}

### 2012-04-24 package/manuscript.Rnw
Just an update.

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
\fixme{Add more coauthors, or change the order of authors if needed} \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

\fixme{Add more corresponding authors.}

@@

\fixme{Add more specific findings while summarizing the results. Connect two
ideas of the differential expression study and transcriptome study.}
A bacterial transcriptome of the primary etiological agent of human dental
caries, \textit{Streptococcus mutans}, is described using deep RNA sequencing.
The technique of RNA-Seq has allowed 
the determination of transcript boundaries for annotated genes and
the identification of previously unknown transcribed regions 
including mRNAs and small noncoding RNAs in the genome.
Differential expression studies on the role of CcpA in 
carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
small RNAs found in our study could be involved in the regulation of CcpA. Our
study provides more precise landscape of transcriptome changes in response to
different carbon sources.
The knowledge that we gained from the current study of \textit{S. mutans}
transcriptome will lead to re-evaluation of the bacterial genome.

@@

% Biological motivations: S. mutans, CCR and CcpA
About one fifth of children of ages ranging from 2 to 19 were reported to have
untreated cavities in the United States \citep{health2010}. Oral bacteria digest
foods containing carbohydrates left on the teeth to convert them to acids, which
dissolve the surface of the teeth and create cavities.  The primary etiological
agent of human dental caries, \textit{Streptococcus mutans}, is one of hundreds
bacterial species that constitute the oral microbial community
\citep{Jenkinson2005}.  Bacteria can selectively use substrates from different
carbon sources via a regulation called carbon catabolite repression (CCR)
\citep{Gorke2008}.  \textit{S.\ mutans} also possesses biological systems for
CCR.  CCR were known to regulates the expression of virulence factors in many
pathogenic bacteria, thereby subjecting a number of genes to CCR.  CcpA was
shown to regulate metabolism and virulence gene expression in 
\textit{S.\ mutans} through CCR in transcriptom study using 
the microarray technique based on DNA hybridization \citep{Abranches2008}.

@@

\subsection*{Experimental results from Burne group}
\fixme{Add some if needed.}

\subsection*{Results from Stanhope group when RNA sequencing preparation}
\fixme{Add some if needed.}

\subsection*{Cluster of RNA samples}

@@

\subsection*{Predictions of small RNAs}

We predicted
\Sexpr{length(colnames(smutans.genes.criteria))-1} putative non-coding RNAs, most of
which were already annotated.  
Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
\Sexpr{length(txRNAz.unannotated)} were in  
unannotated non-coding regions (table \ref{tabUnannotatedRNA}).  Using
\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
shows functional categories enriched with target genes for predicted small RNAs.
We found that a few of the putative small RNAs were structured with stable stems
(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).  
\fixme{We might be able to 
further experimentally verify potentially important small RNAs.}

>>>>

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

@@

\fixme{Add more specific findings while summarizing the results. Connect two
ideas of the differential expression study and transcriptome study.}

A bacterial transcriptome of the primary etiological agent of human dental
caries, \textit{Streptococcus mutans}, is described using deep RNA sequencing.
The technique of RNA-Seq has allowed 
the determination of transcript boundaries for annotated genes and
the identification of previously unknown transcribed regions 
including mRNAs and small noncoding RNAs in the genome.

Differential expression studies on the role of CcpA in 
carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
small RNAs found in our study could be involved in the regulation of CcpA.

Our study provides more precise landscape of transcriptome changes in response to
different carbon sources in \textit{S. mutans}.

The knowledge that we gained from the current study of \textit{S. mutans}
transcriptome will lead to re-evaluation of the bacterial genome.

@@

% Biological motivations: S. mutans, CCR and CcpA
About one fifth of children of ages ranging from 2 to 19 were reported to have
untreated cavities in the United States \citep{health2010}. Oral bacteria digest
foods containing carbohydrates left on the teeth to convert them to acids, which
dissolve the surface of the teeth and create cavities.  The primary etiological
agent of human dental caries, \textit{Streptococcus mutans}, is one of hundreds
bacterial species that constitute the oral microbial community
\citep{Jenkinson2005}.  Bacteria can selectively use substrates from different
carbon sources via a regulation called carbon catabolite repression (CCR)
\citep{Gorke2008}.  \textit{S.\ mutans} also possesses biological systems for
CCR.  CCR were known to regulates the expression of virulence factors in many
pathogenic bacteria, thereby subjecting a number of genes to CCR.  CcpA was
shown to regulate metabolism and virulence gene expression in 
\textit{S.\ mutans} through CCR in transcriptome study using 
the microarray technique based on DNA hybridization \citep{Abranches2008}.

@@

<<eval=FALSE,echo=FALSE,result=hide>>=
# To find Infernal small RNAs expressed in RNA-seq data
grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
smutansData.txInfernal <- smutansData.txAnnotation[1:274]
ol.Infernal <- findOverlaps(smutansData.txInfernal,grTx,type="within")
r <- ranges(grTx[unique(ol.Infernal@subjectHits)])
for (i in seq(length(r))) {
  cat(paste("chr1:",start(r)[i],"-",end(r)[i]),"\n")
}
#chr1: 22518 - 23227 tRNA
#chr1: 23229 - 25062 tRNA
#chr1: 97029 - 99661 mir
#chr1: 99934 - 100544 tRNA
#chr1: 111689 - 118470 mir
#chr1: 152804 - 153769 * STnc490k different position
#chr1: 182583 - 185313 mir
#chr1: 190870 - 191839 tRNA
#chr1: 293153 - 293660 * SRP_bact SRP_euk_arch different position
#chr1: 322211 - 328546 tRNA
#chr1: 413043 - 415691 * T-box
#chr1: 552584 - 553935 mir
#chr1: 626450 - 630612 * P15 (ambiguous) 
#chr1: 660501 - 661809 * L20_leader (strong) LR-PK1 (ambiguous)
#chr1: 666062 - 667070 * TPP (ambiguous but strong)
#chr1: 710297 - 712890 mir
#chr1: 796924 - 798060 * L21_leader (ambiguous)
#chr1: 800164 - 814296 * PyrR (ambiguous)
#chr1: 938383 - 938619 mir
#chr1: 995160 - 998898 snoR72 (ambiguous)
#chr1: 1115782 - 1117618 * Glycin (ambiguous)
#chr1: 1140867 - 1142427 tRNA
#chr1: 1142434 - 1149502 tRNA
#chr1: 1163587 - 1166184 PyrR (ambiguous)
#chr1: 1209827 - 1214087 tRNA
#chr1: 1214631 - 1215039 * L19_leader (ambiguous)
#chr1: 1326825 - 1328325 * CRISPER (ambiguous)
#chr1: 1410520 - 1420619 * His_leader (ambiguous)
#chr1: 1439314 - 1447074 * T-box (ambigous)
#chr1: 1447225 - 1451227 Thr_leader
#chr1: 1487846 - 1488056 mir
#chr1: 1605925 - 1606244 mir
#chr1: 1615311 - 1618362 23S-methyl
#chr1: 1654702 - 1655118 tRNA
#chr1: 1677566 - 1678101 * T-box (ambiguous)
#chr1: 1678104 - 1678469 tRNA
#chr1: 1758499 - 1759936 mir
#chr1: 1783980 - 1787743 mir
#chr1: 1838690 - 1844933 * mir
#chr1: 2023401 - 2026533 tRNA mir

# To find RNAz predicted small RNAs expressed in RNA-seq data
ol <- findOverlaps(smutansData.txRNAz,grTx,type="within")
r <- ranges(grTx[unique(ol@subjectHits)])
for (i in seq(length(r))) {
  cat(paste("chr1:",start(r)[i],"-",end(r)[i]),"\n")
}
#chr1: 301108 - 303712 no
#chr1: 303714 - 304038 * locus100 His_leader (weak)
#chr1: 340223 - 341140 * locus101 
#chr1: 378392 - 382328 
#chr1: 413043 - 415691 * locus103 T-box (strong)
#chr1: 480819 - 482162 * locus104 (ambiguous)
#chr1: 681862 - 684666 * locus107 yybP-ykoY (ambiguous)
#chr1: 742326 - 747734 * locus108 (ambiguous)
#chr1: 908974 - 911426 * locus110 (weak)
#chr1: 1036181 - 1038689 locus2
#chr1: 1209827 - 1214087 locus4 tRNA (false?)
#chr1: 1304269 - 1305793 locus7 (ambiguous)
#chr1: 1678104 - 1678469 locus53 tRNA
#chr1: 1783980 - 1787743 locus69

# Expressed regions without annotations
smutansData.txAllAnnotation <- c(smutansData.txRNAz,smutansData.txAnnotation)
strand(smutansData.txAllAnnotation) <- '*'
smutansData.txAllAnnotation <- gaps(reduce(smutansData.txAllAnnotation))
smutansData.txAllAnnotation <- smutansData.txAllAnnotation[-2:-1]
# Check sum(width(smutansData.txAllAnnotation)) is small or < 10% of 2Mbp.
x <- c()
for (i in seq(length(smutansData.txAllAnnotation))) {
  s <- start(ranges(smutansData.txAllAnnotation))[i]
  e <- end(ranges(smutansData.txAllAnnotation))[i]
  x <- c(x,sum(smutansData.txPileup[s:e])/(e-s+1))
}
elementMetadata(smutansData.txAllAnnotation)[,"score"]<-x
smutansData.txNoAnnotation <- smutansData.txAllAnnotation[order(elementMetadata(smutansData.txAllAnnotation)[,"score"])]
for (i in seq(length(smutansData.txNoAnnotation))) {
  if (abs(start(smutansData.txNoAnnotation)[i] - end(smutansData.txNoAnnotation)[i]) > 50) {
    cat(paste("chr1:",start(smutansData.txNoAnnotation)[i],"-",end(smutansData.txNoAnnotation)[i]),"\n")
  }
}
# This does not work. I think that I have to manually check.
strand(smutansData.txNoAnnotation) <- '+'
x <- elementMetadata(smutansData.txNoAnnotation)[,"score"]
x <- as.integer(x*1000/max(x))
elementMetadata(smutansData.txNoAnnotation)[,"score"] <- x
smutansData.txNoAnnotation <- sort(smutansData.txNoAnnotation)
elementMetadata(smutansData.txNoAnnotation)[,"name"] <- seq(length(smutansData.txNoAnnotation))
export.bed(smutansData.txNoAnnotation,con="b.bed")
x <- c(46, 69, 73, 74, 75, 76, 80, 97, 103, 212, 261, 280, 288, 317, 440, 450, 478,
484 , 628 , 753 , 1011, 1019 , 1037 , 1084, 1097, 1176, 1291, 1444, 1508,
1518, 1600, 1605, 1638, 1645, 1690)
for (i in x) {
  cat(paste("chr1:",start(smutansData.txNoAnnotation)[i],"-",end(smutansData.txNoAnnotation)[i]),"\n")
}

# 46
# 69 * strong
# 73 ambiguous
# 74 very weak
# 75 ambiguous
# 76 ambiguous
# 80 * weak
# 97 *
# 103 *
# 212 * strong
# 261 *
# 280 *
# 288 * one gene?
# 317 * in between tRNAs
# 440 weak
# 450 * strong
# 478 * strong
# 484 * strong
# 628 *
# 753 *
# 1011 * strong
# 1019 * tail of a gene?
# 1037 * strong?
# 1084 * strong
# 1097 * left? PCR amplification error?
# 1176 * start or end of genes?
# 1291 * end of the gene?
# 1444 * 
# 1508 * strong
# 1518 * strong
# 1600 * start of the gene?
# 1605 * new gene in between genes?
# 1638 * start of the gene?
# 1645 * start of the gene?
# 1690 * in between genes?

for (i in seq(length(txRNAz.unannotated))) {
  cat(paste("# ",names(txRNAz.unannotated)[i],"  chr1:",start(txRNAz.unannotated)[i],"-",end(txRNAz.unannotated)[i]),"\n")
}
#  locus99   chr1: 302288 - 302403  no
#  locus100   chr1: 303745 - 303859 did
#  locus101   chr1: 340263 - 340343 did
#  locus102   chr1: 378452 - 378505 good one
#  locus104   chr1: 481458 - 481539 not a good candidate
#  locus106   chr1: 674067 - 674135 maybe one
#  locus108   chr1: 747483 - 747567 not a good one
#  locus110   chr1: 911157 - 911258 not a good one
#  locus2   chr1: 1036231 - 1036283 not a good one
#  locus5   chr1: 1244007 - 1244086 not a good one
#  locus6   chr1: 1297150 - 1297227 not a good one
#  locus7   chr1: 1304792 - 1304926 not a good one
#  locus8   chr1: 1313525 - 1313674 not a good one
#  locus10   chr1: 1564032 - 1564149 not a good one
#  locus69   chr1: 1787575 - 1787687 not a good one
#  locus85   chr1: 1864317 - 1864461 not a good one
@

\subsection*{Small RNAs expressed in RNA-seq}

We compiled 274 predicted RNAs by using cmsearch program from Infernal (v1.0)
with default options for models in RFAM 9.1 \citep{Gardner2009,Nawrocki2009}.
By intersecting the predicted RNAs locations and RNA-seq expressed regions,
we found more evidence for some of the predicted RNAs (table \ref{tab:infernal}).
Several RNAs were shown to be clearly expressed by RNA-seq experiments:
T-box strong, RNaseP\_bact\_a, RNaseP\_bact\_b, L20\_leader, tmRNA, FMN, and 6S.
Other RNAs, TPP and Glycin, were located close to their neighboring genes,
making it hard to discriminate them from the genes.

Additionally, we predicted 
\Sexpr{length(smutansData.txRNAz)}
structural RNAs using RNAz, most of which were overlapped with those from
the Infernal prediction. We found a few of them to be expressed based on RNA-seq
experiments (table \ref{tab:RNAz}).
Region locus100 had tens of short reads mapped on it, which was near
His\_leader predicted by Infernal. 
Although region locus101 was ambiguous in its location with respect to its neighboring
genes, its predicted secondary structure hinted at its potential as a functional
RNA. Region locus103 was expressed clearly compared with neighboring regions; the
region was predicted as T-box by Infernal.
Although region locus107 was ambiguous in its location with respect to its neighboring
genes expression, its good secondary structure and yybP-ykoY predicted by
Infernal hinted at the RNA functional potential of the region.

\subsection*{Unannotated regions expressed in RNA-seq}

We focused 16 regions that had not been annotated either by previously known
genes or by the potentially functional RNA prediction of our current study
(table \ref{tab:unannotated}). Four regions inlcuding novel6,
novel7, novel10, and novel11 showed that their expression levels were not
clearly dicriminated from their flanking genes. 
Located in between SMU.305 and SMU.307 was, for example, novel6, which was
connected in expression level by large number of short reads with SMU.307.
The other 12 regions appeared to show distinct expression levels compared with
flank regions. Region novel1 was situated in between SMU.61 and SMU.63c, which
had different strands.
Although seemingly separated from either of the two genes, it was connected by
links of average about 20 depths of short reads with the either of them.
Regions novel2, novel13, and novel15 appeared to be expressed very weakly.
A few regions were neighboring known RNAs: 
novel3 located in between SMU.97 and SMU.t19,
novel4 located in between SMU.t20 and SMU.107, and
novel6 situated to the right of
SRP\_bact and SRP\_euk\_arch predicted by Infernal.
We also examined regions in between known CDS or tRNAs, which could be
unannotated regions but likely to be in operon with its neighboring genes
(table \ref{tab:inbewteencds}).

% Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
% \Sexpr{length(txRNAz.unannotated)} were in  
% unannotated non-coding regions (table \ref{tabUnannotatedRNA}).

Using
\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
shows functional categories enriched with target genes for predicted small RNAs.

We found that a few of the putative small RNAs were structured with stable stems
(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).

@@

The automatically defined transcripts using site-wise expression values by the
hidden Markov model and mapped short reads displayed using UCSC genome browser
tracks guided us to define transcripts.

tx0391
tx0392

### 2012-04-24 package/supplementary.Rnw
Just an update.

>>>>

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Infernal predicted small RNAs that appeared to be expressed in UA159 Glucose
RNA-seq.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
Position & Description \\
\hline 
chr1: 413043 - 415691 & T-box \\
chr1: 439797 - 440210 & RNaseP\_bact\_a, RNaseP\_bact\_b \\
chr1: 659901 - 661209 & L20\_leader \\
chr1: 666062 - 667070 & TPP (ambiguous) \\
chr1: 1115846 - 1116774 & Glycin (ambiguous) \\
chr1: 1139350 - 1139673 & tmRNA \\
chr1: 1202719 - 1202902 & T-box \\
chr1: 1615122 - 1615307 & FMN \\
chr1: 1927964 - 1928120 & 6S \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:infernal}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf RNAz predicted small RNAs that appeared to be expressed in UA159 Glucose
RNA-seq.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cll}
\hline 
ID & Position & Infernal prediction \\
\hline 
locus100 & chr1: 303714 - 304038 & His\_leader \\
locus101 & chr1: 340223 - 341140 & \\
locus103 & chr1: 413043 - 415691 & T-box \\
% locus104 & chr1: 480819 - 482162 & \\
locus107 & chr1: 681862 - 684666 & yybP-ykoY \\
% locus85 & chr1: 1864317 -1864461 & \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:RNAz}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Expressed regions of unannotated regions in UA159 Glucose
RNA-seq.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cll}
\hline 
ID & Position & Description \\
\hline 
novel1 & chr1: 62546 - 63078 & clear, strong \\
novel2 & chr1: 69293 - 69630 & weak \\
novel3 & chr1: 99553 - 100331 & clear, strong \\
novel4 & chr1: 108631 - 109107 & clear, strong \\
novel5 & chr1: 211452 - 211769 & clear, strong \\
novel6 & chr1: 293238 - 293666 & vague, strong \\
novel7 & chr1: 484053 - 484635 & clear, strong \\
novel8 & chr1: 523427 - 524198 & clear, weak \\
novel9 & chr1: 530356 - 530591 & clear, weak \\
novel10 & chr1: 702752 - 703471 & vague, strong \\
novel11 & chr1: 1202920 - 1203816 & vague \\
novel12 & chr1: 1254303 - 1255083 & clear, strong \\
novel13 & chr1: 1370933 - 1371532 & weak \\
novel14 & chr1: 1757571 - 1758606 & clear, strong \\
novel15 & chr1: 1769455 - 1770330 & clear, weak \\
novel16 & chr1: 76962 - 77575 & SMU.76 and SMU.78 \\
% & chr1: 1291334 - 1291685 & PCR amplification error \\
% & chr1: 1689061 - 1689404 & PCR amplification error \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:unannotated}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Regions expressed in between CDS regions in UA159 Glucose
RNA-seq.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cll}
\hline 
ID & Position & Description \\
\hline 
bt1 & chr1: 68950 - 69214 & between SMU.67 and SMU.68 \\
bt2 & chr1: 268249 - 269322 & between SMU.281 and SMU.283 \\
bt3 & chr1: 301821 - 302287 & between SMU.317 and SMU.318 \\
bt4 & chr1: 328198 - 328359 & between SMU.t34 and SMU.t35 \\
bt5 & chr1: 473362 - 473629 & start of SMU.503c \\
bt6 & chr1: 869137 - 869771 & start of SMU.911c \\
bt7 & chr1: 1176217 - 1176736 & start of SMU.1237c \\
bt8 & chr1: 1185233 - 1185628 & end of SMU.1249c \\
bt9 & chr1: 1534290 - 1534622 & end of SMU.1604.c \\
bt10 & chr1: 1870392 - 1871025 & between SMU.1995c and SMU.1996 \\
bt11 & chr1: 1893131 - 1893647 & start of SMU.2026c \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:inbewteencds}
\end{table}
%-----------------------------------------------------------------------------
